Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-03 paper/paper.tex
Update entrega 1

\setstretch{1.5}
This document was last modified at 2025-11-02 18:14:07 and it was last
rendered at 2025-11-02 18:14:07.

@@

Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}), building on
a critical reassessment of the inferential limits of standard mobility
models, propose a causal framework that reconceptualizes the ``mobility
effect'' as the treatment effect of reaching a destination class, with
impacts heterogeneous by class of origin. The core claim is that typical
mobility hypotheses are fundamentally within-person counterfactual
comparisons rather than between-person contrasts. Defining mobility as
\((M = D - O)\) ---a change from origin \((O)\) to destination \((D)\),
rather than an additive or interactive combination---places the focus
squarely on the Neyman--Rubin problem: for any individual, we observe
the outcome under the realized destination but not under alternative
destinations or under immobility.

Exploiting the temporal ordering of origin, destination and outcome,
Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) frame
mobility as a treatment process that motivates the causal question:
``how would the outcome among people from origin \emph{j} who entered
destination \emph{k} have been different if those people had,
counterfactually, entered destination \emph{k'} instead?'' (p.472). The
corresponding estimand is the conditional causal effect of destination
given origin. It is of particular interest when \emph{k' = j}, i.e.,
immobility. This estimand compares movers' observed outcomes with the
hypothetical outcomes those same individuals would have exhibited had
they remained in their origin class. This formulation provides a
coherent potential-outcomes basis for studying how class transitions
shape individual preferences and attitudes.

@@

\textbf{Positivity}: For all \((j,c)\) in the support of \((O,C)\) and
for each relevant destination \((k)\), the probability of receiving
\(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
has some chance of each comparison destination.

@@

\textbf{Conditional Independence}: Also known as conditional
unconfoundedness or exchangeability, this assumption emphasizes that,
conditional on origin \(O\) and pre-treatment covariates \(C\) (e.g.,
parental education, ethnicity, cohort, early-life factors), assignment
to \(D\) is as good as random. This underlies IPW and regression
adjustment, which seek exchangeability between mobile and immobile
individuals.

@@

Figure~\ref{fig-dag} depicts the causal model guiding identification. In
this DAG, \(C\) represents pre-treatment confounders (e.g., family
resources, household composition, and ascriptive status) affecting both
origin and destination, and \(U\) denotes unobserved factors influencing
\(C\). Class origin \(O\) is determined by \(C\) and, in turn, affects
destination \(D\). The outcome \(Y\) (preferences for commodification)
is caused by \(D\) and can also be affected by \(C\). There is no direct
arrow from \(O\) to \(Y\), reflecting the assumption that the effect of
origin on the outcome operates entirely through background factors
\((C)\) and destination \((D)\). Accordingly, conditioning on \((O,C)\)
blocks the relevant back-door paths from \(D\) to \(Y\), specifically
\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
adjustment set.

@@

In this study, I focus on the specific case where the counterfactual
destination corresponds to class immobility \((k' = j)\), that is, how
the outcome among people from origin \emph{k} who entered destination
\emph{k} would have bee different if those people had, counterfactually,
remained in origin \emph{j} instead (inmobility). Accordingly, the
estimand becomes:

@@

Because respondent occupation is not measured in every wave, I restrict
the analysis to the 2016, 2018, and 2023 waves. Parental occupation was
collected only in 2023 as open-text labels. I coded these texts to the
ISCO-08 two-digit scheme using the National Institute of Statistics of
Chile's automated coding API and retained cases with high-confidence
matches (\textgreater95\%). Treating parental occupation as a
time-invariant origin attribute, I carried these codes back to 2016 and
2018. After listwise deletion and restricting to key variables
(respondent occupation and the outcome measure), the final analytic
sample comprises \emph{N} = 3,435 observations nested within \emph{N} =
1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with
the study design, estimation proceeds within trajectory-specific
subsamples (e.g., low→low; low→middle), retaining movers and matched
non-movers from the same origin class. Then, I apply weights to
construct the corresponding counterfactual control distributions for
each trajectory. As Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}) did, I use the three waves in
the estimations that follow, allowing for correlation between the
observations for each individual in calculating the standard errors
(i.e.~clustering on the personal identifier).

@@

The outcome variable measures preferences regarding the commodification
of pensions, operationalized with a single item addressing the degree of
public support for market-based access to old-age pension benefits.
Respondents were asked: ``Is it fair in Chile that people with higher
incomes have better pensions than people with lower incomes?'' and were
asked to indicate their level of agreement on a five-point Likert scale
ranging from 1 (``strongly disagree'') to 5 (``strongly agree''). For
estimation and interpretability, I dichotomized the item so that values
4-5 indicate agreement with market-based access to old-age pensions
(coded as 1), while values 1-3 indicate disagreement (coded as 0). This
binary coding treats neutrality as substantially closer to non-support
than to support and produces a clear contrast between respondents who
support market-based differentiation in pensions and those who do not. I
use this item for two main reasons: on the one hand, to be able to make
comparisons with existing literature
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}) and, on the
other hand, because it captures two fundamental dimensions of
market-based welfare distribution: the role of economic resources as a
key determinant of outcomes and the framing of services as as tradable
commodities that can be bought and sold based on ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

@@

After balancing, I construct stabilized IPW-ATT, setting treated weights
to 1 and assigning ebal-derived weights to controls, thus aligning the
counterfactual distribution of non-movers with the covariate profile of
movers in each origin stratum (see \hyperref[anexo]{Supplementary
Material} for the evaluation of balance). These weights are then used in
the outcome stage to estimate the causal effect of mobility on
preferences for the commodification of pensions.

@@

Because the dependent variable---the preference for pension
commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
linear probability models (WLS) using stabilized inverse probability
weights (IPW--ATT) obtained via entropy balancing. Although OLS
estimator is often questioned with binary outcomes, it's unbiased. It
consistently estimates the conditional expectation \(E(Y|X)\) under
standard exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity
addressed by robust or clustered errors
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
ch.~7.5). WLS further improves efficiency under heteroskedasticity or
unequal sampling probabilities
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
ch.~8.4). Following Gelman, Hill, and Vehtari
(\citeproc{ref-gelman_regression_2020}{2020, pp. 270--272}),
IPW-weighted regression transparently targets the ATT by recreating the
counterfactual distribution. Linear models are preferred here because
coefficients are average probability differences; nonlinear models
(logit/probit) complicate weighting and interpretation
(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020}, ch.~13).

@@

The specification is doubly robust: entropy-balancing weights are
combined with covariate adjustment so consistency holds if either the
weighting model or the outcome model is correctly specified, while also
improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). I
report both IPW-only and doubly robust specifications to document
robustness. In addition, I assess causal heterogeneity by interacting
the mobility treatment with (i) meritocratic perceptions and (ii) survey
wave, reporting simple slopes (origin-specific ATTs) by moderator
levels. To evaluate sensitivity to unmeasured confounding, I compute
E-values for point estimates and confidence limits
(\citeproc{ref-ding_sensitivity_2016}{Ding \& VanderWeele, 2016}). To
address multiplicity across the six mobility trajectories, I also report
Holm-adjusted p-values alongside 95\% confidence intervals.

@@

Figure~\ref{fig-reg} reports the doubly robust estimates of the effect
of intergenerational mobility on preferences for pension
commodification, obtained from weighted models combining stabilized
IPW--ATT with covariate adjustment and wave fixed effects. This
specification ensures consistency if either the weighting or the outcome
model is correctly specified, while improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). Results
are virtually identical to baseline models without covariates (see
\hyperref[anexo]{Supplementary Material} for complete models), with no
substantive change in the size or significance of the mobility
coefficients.

@@

Results reveal a clear directional asymmetry between upward and downward
mobility trajectories. Among upwardly mobile respondents, only the
Middle→High trajectory exhibits a significant positive effect on support
for pension commodification (\(\beta\) = 0.09, 95\% CI {[}0.02,
0.16{]}). Interpreted as an origin-specific ATT, this estimate
compares---for individuals who actually moved from the middle to the
high class---their observed endorsement with the counterfactual
endorsement they would have expressed had they remained immobile in the
middle class. Holding all other predictors constant, the point estimate
implies a 9 percentage point higher probability of endorsing
market-based access to pension benefits for Middle→High movers relative
to their own immobility counterfactual, statistically significant at the
95\% confidence level. In contrast, two downward trajectories:
High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
effects, suggesting that individuals who experience downward mobility
from higher-class origins express weaker preferences for market
provision. The remaining pathways (Low→Middle, Low→High, Middle→Low)
display no significant differences relative to their non-mobile
counterparts. These patterns indicate that upward movement within the
upper segment (Middle→High) reinforces pro-commodification attitudes,
whereas downward movement from privileged origins reduces them.

@@

To probe potential mechanisms underlying the mobility effects, I
estimate origin-specific ATT models that interact the mobility treatment
with high meritocratic perceptions (ref. = low), using stabilized
IPW--ATT weights, covariate adjustment, and wave fixed effects. I also
estimate an analogous specification interacting mobility with survey
waves (ref. = 2016) to assess temporal heterogeneity in the treatment
effect (see \hyperref[anexo]{Supplementary Material} for complete
models).

@@

\section{Supplementary material}\label{anexo}

>>>>

\setstretch{1.5}
This document was last modified at 2025-11-03 01:00:05 and it was last
rendered at 2025-11-03 01:00:05.

@@

Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}), building on
a critical reassessment of the inferential limits of standard mobility
models, propose a causal framework that reconceptualizes the ``mobility
effect'' as the treatment effect of reaching a destination class, with
impacts heterogeneous by class of origin. The core claim is that typical
mobility hypotheses are fundamentally within-person counterfactual
comparisons rather than between-person contrasts. Defining mobility as
\(M = D - O\) ---a change from origin \((O)\) to destination \((D)\),
rather than an additive or interactive combination---places the focus
squarely on the Neyman--Rubin problem: for any individual, we observe
the result in the social position they currently occupy, but not in the
situation of other alternative destinations or immobility.

Exploiting the temporal ordering of origin, destination and outcome,
Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) frame
mobility as a treatment process that motivates the causal question:
``how would the outcome among people from origin \emph{j} who entered
destination \emph{k} have been different if those people had,
counterfactually, entered destination \emph{k'} instead?'' (p.472). The
corresponding estimand is the conditional causal effect of destination
given origin. It is of particular interest when \emph{k' = j}, i.e.,
immobility. This estimand compares movers' observed outcomes with the
hypothetical outcomes those same individuals would have exhibited had
they remained in their origin class. This formulation provides a
coherent potential-outcomes basis for studying how social mobility shape
individual preferences and attitudes.

@@

\textbf{Positivity}: For all \((j,c)\) in the support of \((O,C)\) and
for each relevant destination \((k)\), the probability of receiving
\(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
has non-zero chance of each comparison destination.

@@

\textbf{Conditional Independence}: Also known as conditional
unconfoundedness or exchangeability, this assumption emphasizes that,
conditional on origin \(O\) and pre-treatment covariates \(C\) (e.g.,
parental education, ethnicity, cohort, early-life factors), assignment
to \(D\) is as good as random. This underlies IPW and regression
adjustment, which seek exchangeability between mobile (treated) and
immobile individuals (controls).

@@

In fact, establishing the causal relationship between social mobility
and subjective outcomes, such as preferences, poses several challenges
for causal inference. Below, I use directed acyclic graphs (DAGs) to
illustrate some of these challenges and evaluate possible strategies for
identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
the causal model guiding identification. In this model, a person's
preference (\(Y\)) is directly caused by her class destination (\(D\))
and can also be affected by different attributes determined in childhood
or earlier (\(C\)), such as family resources, household composition, and
ascriptive status (pre-treatment confounders). Class origin (\(O\)) is
determined by those origin-attributes (\(C\)) and, in turn, affects
destination (\(D\)). In addition, denotes unobserved factors (\(U\)) can
also influence origin-attributes (\(C\)). There is no direct arrow from
\(O\) to \(Y\), reflecting the assumption that the effect of origin on
the outcome operates entirely through background factors \((C)\) and
destination \((D)\). Accordingly, conditioning on \((O,C)\) could blocks
the relevant back-door paths from \(D\) to \(Y\), specifically
\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
adjustment set.

@@

In this study, I focus on the specific case where the counterfactual
destination corresponds to class immobility \((k' = j)\), that is, how
the outcome among people from origin \emph{j} who entered destination
\emph{k} would have bee different if those people had, counterfactually,
remained in origin \emph{j} instead (inmobility). Accordingly, the
estimand becomes:

@@

Because respondent occupation is not measured in every wave, I restrict
the analysis to the 2016, 2018, and 2023 waves. Parental occupation was
collected only in 2023 as open-text labels. I coded these texts to the
ISCO-08 two-digit scheme using the National Institute of Statistics of
Chile's automated coding API and retained cases with high-confidence
matches (\textgreater95\%). Treating parental occupation as a
time-invariant origin attribute, I carried these codes back to 2016 and
2018. After listwise deletion and restricting to key variables
(respondent occupation and the outcome measure), the final analytic
sample comprises \emph{N} = 3,435 observations nested within \emph{N} =
1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with
the study design, estimation proceeds within trajectory-specific
subsamples (e.g., low→low; low→middle), retaining movers and matched
non-movers from the same origin class. Then, I apply weights to
construct the corresponding counterfactual control sample for each
trajectory. Following Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}), I use the three waves in the
estimations that follow, allowing for correlation between the
observations for each individual in calculating the standard errors
(i.e.~clustering on the personal identifier).

@@

The outcome variable measures preferences regarding the commodification
of pensions, operationalized with a single item addressing how strongly
individuals justify conditioning access to old-age pension benefits
based on individual income. Respondents were asked: ``Is it fair in
Chile that people with higher incomes have better pensions than people
with lower incomes?'' and were asked to indicate their level of
agreement on a five-point Likert scale ranging from 1 (``strongly
disagree'') to 5 (``strongly agree''). For estimation and
interpretability, I dichotomized the item so that values 4-5 indicate
agreement with market-based access to old-age pensions (coded as 1),
while values 1-3 indicate disagreement (coded as 0). This binary coding
treats neutrality as substantially closer to non-support than to support
and produces a clear contrast between respondents who support
market-based differentiation in pensions and those who do not. I use
this item for two main reasons: on the one hand, to be able to make
comparisons with existing literature
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}) and, on the
other hand, because it captures two fundamental dimensions of
market-based welfare distribution: the role of economic resources as a
key determinant of outcomes and the framing of services as as tradable
commodities that can be bought and sold based on ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

@@

After balancing, I construct stabilized IPW, setting treated weights to
1 and assigning derived weights to controls, thus aligning the
counterfactual distribution of non-movers with the covariate profile of
movers in each origin stratum (see \hyperref[anexo]{Supplementary
Material} for the evaluation of balance). These weights are then used in
the outcome stage to estimate the causal effect of mobility on
preferences for the commodification of pensions.

@@

Because the dependent variable---the preference for pension
commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
linear probability models (WLS) using stabilized inverse probability
weights (IPW) obtained via entropy balancing. Although OLS estimator is
often questioned with binary outcomes, it's unbiased. It consistently
estimates the conditional expectation \(E(Y|X)\) under standard
exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity addressed
by robust or clustered errors
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
ch.~7.5). WLS further improves efficiency under heteroskedasticity or
unequal sampling probabilities
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
ch.~8.4). Following Gelman, Hill, and Vehtari
(\citeproc{ref-gelman_regression_2020}{2020, pp. 270--272}),
IPW-weighted regression transparently targets the ATT by recreating the
counterfactual distribution. Linear models are preferred here because
coefficients are average probability differences; nonlinear models
(logit/probit) complicate weighting and interpretation
(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020}, ch.~13).

@@

The specification is doubly robust: entropy-balancing weights are
combined with covariate adjustment so consistency holds if either the
weighting model or the outcome model is correctly specified, while also
improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
results are virtually identical to baseline models without covariates
(see \hyperref[anexo]{Supplementary Material} for complete models), with
no substantive changes in the mobility coefficients. In addition, I
assess causal heterogeneity by interacting the mobility treatment with
(i) meritocratic perceptions and (ii) survey wave, reporting simple
slopes (origin-specific ATTs) by moderator levels. To evaluate
sensitivity to unmeasured confounding, I compute E-values for point
estimates and confidence limits
(\citeproc{ref-ding_sensitivity_2016}{Ding \& VanderWeele, 2016}). To
address multiplicity across the six mobility trajectories, I also report
Holm-adjusted p-values alongside 95\% confidence intervals.

@@

Figure~\ref{fig-reg} reports the doubly robust estimates of the effect
of intergenerational mobility on preferences for pension
commodification, obtained from weighted models combining stabilized IPW
with covariate adjustment and wave fixed effects. Results reveal a clear
directional asymmetry between upward and downward mobility trajectories.
Among upwardly mobile respondents, only the Middle→High trajectory
exhibits a significant positive effect on support for pension
commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
Interpreted as an origin-specific ATT, this estimate compares---for
individuals who actually moved from the middle to the high class---their
observed endorsement with the counterfactual endorsement they would have
expressed had they remained immobile in the middle class. Holding all
other predictors constant, the point estimate implies a 9 percentage
point higher probability of endorsing market-based access to pension
benefits for Middle→High movers relative to their own immobility
counterfactual, statistically significant at the 95\% confidence level.
In contrast, two downward trajectories: High→Middle (\(\beta\) = −0.07,
CI {[}−0.15, −0.00{]}) and High→Low (\(\beta\) = −0.10, CI {[}−0.18,
−0.02{]}), show significant negative effects, suggesting that
individuals who experience downward mobility from higher-class origins
express weaker preferences for market provision. The remaining pathways
(Low→Middle, Low→High, Middle→Low) display no significant differences
relative to their non-mobile counterparts. These patterns indicate that
upward movement within the upper segment (Middle→High) reinforces
pro-commodification attitudes, whereas downward movement from privileged
origins reduces them.

@@

@@

To probe potential mechanisms underlying the mobility effects, I
estimate origin-specific ATT models that interact the mobility treatment
with high meritocratic perceptions (ref. = low), using stabilized IPW
weights, covariate adjustment, and wave fixed effects. I also estimate
an analogous specification interacting mobility with survey waves (ref.
= 2016) to assess temporal heterogeneity in the treatment effect (see
\hyperref[anexo]{Supplementary Material} for complete models).

@@
